phoeton: Software for Objective Probabilistic Foehn Classification

### 2018-12-01 inst/draft/draft.tex
Fixed some of the serious typos, still very early note file and not really a draft

Introduction why this is so great. Georg might like to contribute
here. Just to leave/include some references:

\begin{itemize}
    \item finite mixture model \citep{leisch2004,gruen2008}
    \item automatic probabilistic foehn classification \cite{plavcan2014}
    \item IWLS/IRLS \citep{mccullagh1989}
\end{itemize}

% -------------------------------------------------------------------
% SECTION: METHODS
% -------------------------------------------------------------------
\section{Methods}

>>>>

Introduction why this is so great. Georg might like to contribute
here. Just to leave/include some references:
\begin{itemize}
    \item finite mixture model \citep{leisch2004,gruen2008}
    \item automatic probabilistic foehn classification \cite{plavcan2014}
    \item IWLS/IRLS \citep{mccullagh1989}
    \item Community foehn classification experiment \cite{mayr2018unpublished}
\end{itemize}

I would go for something like:

Gap winds are typically characterized by strong increase of wind speed with a rapid
rise in temperature and a drop in relative humidity during the onset. Such winds can be
found in various regions on our globe and bear different names for specific
regions such as the Santa-Anna winds in California which became infamous due to
their destructive effect in combination with wild fires, the Chinook winds in
the Rocky Mountain area, Mistral and Bora in south France and Croatia, or the
Foehn winds in the European Alps.  Over the past decades \textit{foehn} wind
has become a synonym for such gap flows and will be used in this article.
However, the method as presented is applicable for all gap flow winds
independent from their geographical location.

The conceptual model of foehn winds is quite well known
\citep{armi2007,mayr2008,armi2011}, however, foehn cannot measured directly.
Historically, foehn was often diagnosed manually by looking at available
meteorological records such as wind speed and direction, time series of
temperature and humidity, the conditions of the sky (e.g., occurrence of
lenticularis clouds in the region), and observations upstream and downstream of
the location of interest.  Some of the longest records of foehn observations
can be found in Switzerland (Altdorf \note{i'm just guessing!}) reaching back
to 1850 (\note{just guessing!}) for a hand full of stations.  These records are
typically based on manual classification made by experienced weather observers,
weather enthusiasts, and meteorologists.  This task is very time consuming and
is only done for some selected sites, typically sites with long historical
records. To be able to extend the diagnosis and to perform foehn classification
on a large number of meteorological sites, computer aimed classification is
needed.

Few attempts to automatically classify foehn winds have been proposed over the
past years such as \note{cite this and that}. These methods are mostly binary
(either \textit{yes} or \textit{no}) and require manual tuning to specify the
hyper-parameters, e.g., thresholds used to classify a specific observation as
foehn. \cite{plavcan2014} proposed a new automated probabilistic foehn
classification method based on finite mixture models, a statistical
regression-based model for automatic foehn classification. In a recent study of
\cite{mayr2018unpublished} this new method has been tested against existing
classification methods and manual foehn classification made by students and
experts to proof the diagnostic power of the new approach.

Overall, the automatic finite mixture model based foehn classification
\citep{plavcan2014} is able to compete with the human classification
\citep{mayr2018unpublished} and has the advantages that it is (i) reproducible,
and (ii) can easily be applied to large volumes of data. This allows one to
efficiently perform foehn classification for a wide range of different sites.

The article by \citep{plavcan2014} introduced the method using a powerful
package called \code{FlexMix} \citep{leisch2004,gruen2007,gruen2008} available
for the statistical software \textit{R}.  However, only little guidance was
given by \citep{plavcan2014} on how to apply this method to own data and an
implementation of the method outside the \textit{R} programming language was
not available until today.  Thus, the aim of this article is to introduce the
algorithm behind the automated probabilistic foehn classification
\citep{plavcan2014} in more detail and introducing new open-source software for
\textit{R} and \textit{Python} which provides the required methods to perform
the classification.  Furthermore, the software comes with a set of examples and
diagnostic tools for this specific task which allows everyone to use the method
on their own.

Section~\ref{sec:methods} describes the method itself followed by
Section~\ref{sec:software}  introducing the software with a set of
examples. Section~\dots \note{case study? Required for this article?}

% -------------------------------------------------------------------
% SECTION: METHODS
% -------------------------------------------------------------------
\section{Methods}\label{sec:methods}
